Algebraic Groups\vspace{-1ex}

### 2019-11-25 Algebraic Geometry/Algebraic-Geometry.tex
update

	$f$ is \textbf{dominant} if there exists an open $V\subseteq Y$ that is contained in $\Im \tilde f$ for some representative $\tilde f$ of $f$.
\end{defn}
\begin{prob}
	Show that if $f:X\dashrightarrow Y$ is birational then there exist nonempty opens $U$ and $V$ in $X$ and $Y$, respectively such that $f$ induces an isomorphism $U\xrightarrow{\sim} V$ of varieties.
\end{prob}
\begin{defn}
	An irreducible variety is \textbf{rational} if it is birational to $\P n$ for some $n$.
\end{defn}
\begin{rmk}
	Roughly speaking, we think that ``most of $X$ can be uniquely described by $n$ algebraic parameters.'' Try to look at some examples! The points where this doesn't make sense is jsut the 
\end{rmk}
\begin{prob}
	I am repeating some of the old problems to refresh their importance: try figuring out how a quadric projects from $\P2$ to $\P1$ by the usual projection map. There is actually an isomorphism!
\end{prob}
We've been sniffing around the idea that any non-degenerate quadric $Q\subseteq \P n$ is rational. What about cublics? WHich are rational? The cubic curves (defined by cubic equations in $\P2$)
are genus one and NOT rational. The cubic surfaces (those in $\P3$) are apparently.

>>>>

	$f$ is \textbf{dominant} if there exists an open $V\subseteq Y$ that is contained in $\Im \tilde f$ for some representative $\tilde f$ of $f$.
\end{defn}
\begin{prob}
	Show that if $f:X\dashrightarrow Y$ is birational then there exist nonempty opens $U$ and $V$ in $X$ and $Y$, respectively such that $f$ induces an isomorphism $U\xrightarrow{\sim} V$ of varieties.
\end{prob}
\begin{defn}
	An irreducible variety is \textbf{rational} if it is birational to $\P n$ for some $n$.
\end{defn}
\begin{rmk}
	Roughly speaking, we think that ``most of $X$ can be uniquely described by $n$ algebraic parameters.'' Try to look at some examples! The points where this doesn't make sense is jsut the 
\end{rmk}
\begin{prob}
	I am repeating some of the old problems to refresh their importance: try figuring out how a quadric projects from $\P2$ to $\P1$ by the usual projection map. There is actually an isomorphism!
\end{prob}
We've been sniffing around the idea that any non-degenerate quadric $Q\subseteq \P n$ is rational. What about cublics? Which are rational? The cubic curves (defined by cubic equations in $\P2$)
are genus one and NOT rational. The cubic surfaces (those in $\P3$) are apparently.

\section{November 25th, 2019}
Last time we talked about quadrics and cubics and stuff! We asked whether smooth quadrics are all rational (birational to $\P n$). We also asked when cubics are rational.

We saw that when $Q\subseteq \P2$ that $Q\cong \P1.$ In general we can see that these are rational! This involves taking any point on the quadric 
and then project onto the other one. But the important thing to notice here is that this required us to pick a point and we break the symmetry of having 
two points in each fiber.

What about cubics? Here the center of a projection $\P3\dashrightarrow\P1$ is a \textit{line}. Let $S=Z(\text{cubic form})$. Then the fiber over $p\in\P1$ is 
a plane: $\P2\hookrightarrow\P3$ via $(x_0:x_1:x_2)\mapsto(L_0:L_1:L_2:L_3)$ where the $L_i$ are linear forms in the $x_i$. Then if $F$ is our cubic form,
we can consider the slice by this plane: $F(L_0,\dots,L_3)=G(x_0,\dots,x_3)$. For next time: what if the center $l\subseteq S$ (the center of projection is in $S$). One question:
does the intersection $G$ of a plane containing such a line factor as a product of a quadric and a linear factor?

### 2019-11-25 Algebraic Groups/Algebraic-Groups.tex
update

>>>>

\section{November 22, 2019}
From last time (which I missed): we had
\begin{prop}
	Let $G$ be a linear algebraic group over $k$. Then the following are equivalent:
	\begin{itemize}
		\item $G$ is diagonalizable (i.e. $G=D(A)$ for some finitely generated abelian group $A$)
		\item There exists a faithful representation $G\to \GL(V)\cong \GL_n$ which maps into the diagonal matrices.
		\item $\Gamma(G)$ is spanned by grouplike elements.
		\item (Discussed below) All $G$ representations are direct sums of one-dimensional ones.
	\end{itemize}
\end{prop}
Earlier this quarter we saw:
\begin{prop}
	If $G=D(A)$, then any $G$ representation $V$ can be decomposed as 
	\[V=\oplus_{a\in A V_a}\]
	where 
	\[V_a=\{v\in V|\sigma^\ast(v)=t^a\otimes v\}\]
\end{prop}
\begin{rmk}
	Here we used the fact (shown yesterday?) that the characters of an abelian group are in bijection with elements.
\end{rmk}
Note that if $k_a$ is the one-dimensional representation $k_a\to\Gamma(G)\otimes k_a$ via $1\mapsto t^1\otimes 1$. One can 
quickly show that $V_a\cong k_a$, so we get that every representation decomposes as a direct sum of one-dimensional subgroups. In particular, 
$G=D(A)$ are \textit{linearly reductive}.

\subsection{Other properties of diagonalizable groups}
First of all, these groups are commutative. That is, $Hom_\AlgGrp(G,\Gm)\cong A$.

Next, we claim that $\Hom_\AlgGrp(G,\Ga)=0$. To see this, assume that $f:G\to \Ga$ is such a map and consider the 
induced map $f^\ast:k[x]\to k[A]$. Then following through the commutative diagram for comultiplication, we get
\[f^\ast(x)\otimes f^\ast(x)=f^\ast(x)\otimes 1+1\otimes f^\ast(x)\]
by the diagonalizability of $G$ and the fact that $\Ga$ is ``infinitesimal'' (I forget the actual term from Hopf algebras.)

\begin{defn}
	A linear algebraic group $G$ over $k$ is \textbf{of multiplicative type} if there exists a finite extension $k'/k$ such that $G_{k}=G\times_{\Spec k}\Spec k'$ is diagonaliable.
\end{defn}
\begin{rmk}
	This is equivalent to $G_{\bar k}$ being diagonalizable. One direction is obvious. For the other direction, Supposed we ahve a faithful 
	representation $\bar V$ of $\bar G=G_{\bar k}$ by using some descent ideas.
\end{rmk}

\begin{thm}
	Let $G$ be an algebraic group over $k$. Then the following are equivalent:
	\begin{itemize}
		\item $G$ is of multiplicative type.
		\item $G$ is commutative and $\Hom(G,\Ga)=0$.
		\item $G$ is commutative and $\Gamma(G)$ is co\'etale.
		\item $G_{k^s}$ is diagonalizable. That is, there exists $k'/k$ that is finite and seperable such that $G_{k'}$ is seperable.
	\end{itemize}
\end{thm}
What does that mean?
\begin{defn}
	Let $(C,\Delta,\varepsilon)$ be a coalgebra. Let $A=C^\vee$ be the $k$-linear dual. This gives us (using that $C^\vee\otimes C^\vee\hookrightarrow (C\otimes C)^\vee$)
	an algebra. Then 
	\begin{itemize}
		\item If $C$ is finite dimensional, we say that $C$ is \textbf{co\'etale} if $A$ is commutative and if $A=C^\vee$ is \'etale. That is 
		$A=\prod_{i=1}^n k_i'$ where $k_i'/k$ are all finite and seperable extensions.
		\item If $C$ is infinite dimensional, then $C$ is \textbf{co\'etale} if it is a union of finite dimensional co\'etales.
	\end{itemize}
\end{defn}
\begin{ex}
	$G=\Gm$, $\Gamma(G)=k[t]_t$. Then let $C=k\langle t^d\rangle\subseteq \Gamma(G)$. Then $A=C^\vee$ is just $k$, so it is co\'etale.
	Furthermore $k\langle t^{-d},\dots,t,\dots,t^d\rangle$ has that $C^\vee\cong \prod k$, so is as well.

	Finally this gives us that $\Gamma(\Gm)$ is co\'etale since it is a union of these things.
\end{ex}
As an aside, if $V\to \Gamma(G)\otimes V$ is a representation, then $\Im(V\otimes V^\vee\to \Gamma(G))$ is a coalgebra.

\section{November 25th, 2019}
Today we finish our discussion of group schemes of multiplicative type. There was one implication we skipped in Friday's discussion:
that if $G$ is commutative then $\Hom(G,\Ga)=0$ implies that $\Gamma(G)$ is co\'etale.
\subsection{Continuing from last time}
\begin{prf}[cont.]
	Assume that $k=\bar k$ and let $C\subseteq\Gamma(G)$ be a subcoalgebra of finite dimension. Let $A=C^\vee$ and $\langle\cdot,\cdot\rangle$ be 
	the natural pairing $C\times A\to k$. If $A$ is not \'etale, then since $A$ is finite dimensional,
	\[A=A_1\times A_n\]
	where the $A_i$ are local Artinian $k$-algebras.

	So there exists a projection map $\pi:A\to k[\varepsilon]/\varepsilon^2\cong k\oplus k\varepsilon$ that sends 
	\[x\mapsto \langle x,c\rangle+\varepsilon\langle x,d\rangle\]
	for some $c,d\in C$.
	Then a subclaim is to prove that
	\begin{itemize}
		\item $\Delta(c)=c\otimes c$
		\item $\Delta(d)=c\otimes d+d\otimes c$
		\item $\varepsilon(c)=1$
	\end{itemize}
	Then by also showing that $c$ is invertible, one can compute 
	\[\Delta(dc^{-1})=1\otimes c^{-1}d+c^{-1}d\otimes 1\]
	which then gives us a (nontrivial!) map $G\to \Ga$, which contradicts the assumption.
\end{prf}
\begin{ex}[Non co\'etale example]
	Consider $\Ga=\Spec k[x]$ with $\delta(x)=x\otimes 1+1\otimes x$. Then consider th subcoalgebra spanned by $\langle 1,\dots,x^{n-1}\rangle$
	and the dual algebra $A=C^\vee$ spanned by $e_i$. One can compute that 
	\[e_i\cdot e_j=\binom{i+j}{i}e_{i+j}\]
	if $i+j\le n-1$ and it is zero otherwise. But then in particular $e_{n-1}^2=0$, so $A$ isn't even reduced!

	One could ask whether this still holds for any choice of subcoalgebra, but we aren't doing that.
\end{ex}
\begin{cor}
	Let $G$ be a linear algebraic group over $k$. Assume that $G$ is commutative (and maybe smooth). Then $G$ is of multiplicative type 
	if and only if $G(\bar k)$ consists purely of semisimple elements.
\end{cor}
\begin{cor}
	Let 
	\[1\to G'\to G\to G''\to 1\]
	be a short exact sequence. If $G$ is commutative, then $G$ is of multiplciative type if and only if both $G'$ and $G''$ are.
\end{cor}
\begin{rmk}
	That subgroups and quotients of multiplicative type group schemes are multiplicative type is not hard to see. To see this, apply $\Hom(-,\Ga)$ to the sequence and notice that quotients and subgroups 
	of commutative groups are commutative.

	The nontrivial direction is that the collection of multiplicative type groups is closed under extension. 
\end{rmk}
